Characterizing dynamical transitions by statistical complexity measures based on ordinal pattern transition networks

### 2020-11-26 manuscript_Huang.tex
Changes made by Donner

\begin{abstract}
Complex network approaches have been recently emerging as novel and complementary concepts of nonlinear time series analysis which are able to unveil many features that are hidden to more traditional analysis methods. In this work, we focus on one particular approach: the application of ordinal pattern transition networks for characterizing time series data. More specifically, we generalize some traditional estimators of statistical complexity measures (SCMs), which explicitly disclose heterogeneous frequencies of ordinal pattern transitions in the time series networks. To demonstrate the usefulness of the estimators, we employ SCMs to characterize different dynamical transitions in the logistic map as a paradigmatic model system, as well as real-world time series of fluid experiments and ECG data. The obtained results for both, artificial and experimental data demonstrate that the consideration of transition frequencies between different ordinal patterns leads to dynamically meaningful estimates of SCMs, which provide prospective tools for the analysis of time series with finite length. 
\end{abstract}

@@

\begin{quotation}
In the recent decade, traditional tools of nonlinear time series analysis have been undergoing fast developments benefiting from concepts from complex network theory. Along this line of research topics, ordinal pattern transition networks have been expanding the established concept of permutation entropy and provide complementary insights into the dynamical organization underlying time series data. Traditionally, permutations based on order patterns are simple and easy to implement concepts providing estimators of statistical complexity measures (SCMs), which however rely on static pattern frequencies only. Yet, much additional information can be exploited by including order pattern transitions into the definitions of SCMs -- an idea that however has not been widely developed and applied so far. In this work, we generalize existing estimators of SCMs by means of ordinal pattern transition networks which take into account the pattern transition properties explicitly. The usefulness of our generalizations of existing SCMs is demonstrated by using time series of both, model and experimental data. 
\end{quotation}

\section{Introduction}
In the recent decade, complex network based approaches have emerged as prominent tools for nonlinear time series analysis \cite{Kantz97,Sprott2003}, which have found many interesting applications to experimental data from various origins \cite{ZouPR2018}. In this line of research, there exist several major methods that are of particular importance, including recurrence networks \cite{MarwanPLA2009,Donner2010a}, visibility graphs \cite{Lacasa2008,Nunez2012}, transition networks \cite{Nicolis2005,MichaelChaos2015} and cycle networks \cite{ZhangPRL2006}. Depending on the particular questions at hand, each of these complex network approaches exhibit various variants that can be employed for different application purposes. For instance, we may either construct a single network representing a univariate time series, or interacting, multiplex or multi-layer networks for coupled time series. For a recent systematic review, see Ref.~\cite{ZouPR2018}. In this work, we focus on constructing ordinal pattern transition networks (OPTNs) for time series data, which present the advantages of easy implementation and a wide range of existing application to data from different origins, including correlated stochastic processes, neurophysiological (EEG) and human cardiac activity (ECG) series \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019,Subramaniyam2020}.

@@

More specifically, given a one-dimensional time series $\{ x_i\}_{i=1, \cdots, L}$ comprising $L$ observations from a dynamical system, we first reconstruct the corresponding phase space by time delay embedding $\vec{x}_i = [x_i, x_{i+\tau}, \ldots, x_{i+(D-1)\tau}]$ with an embedding dimension $D$ \cite{Takens1981,Kantz97}. Next, we represent each embedding vector $\vec{x}(i)$ by the corresponding rank order of its components, which is encoded into a symbol $\pi_{i}$. Hence, when sliding windows from $i=1$ to $N = L - (D - 1)\tau$ in the embedding space, a symbolic representation of the trajectory $\{\pi_i\}$ is produced. It is possible to derive information about the dynamics of the underlying system by assessing the probabilities of occurrence of different ordinal patterns. For example, for time series resulting from deterministic dynamical systems, certain patterns may not occur at all. More generally, we obtain an empirical probability distribution $P$ whose elements $p_{\pi_{i}}$ are the frequencies associated to the different patterns $\pi_{i}$, $i = 1, \cdots, D!$. Clearly, $P$ provides a significant and feasible way to estimate a characteristic probability distribution function for a given time series, which plays a crucial role in extracting statistical complexity information on the underlying dynamical systems \cite{BandtPRL2002,AmigoPTRSA2014}.

The conceptually relatively simple computation of $P$ based on ordinal patterns for the underlying system prompts to another successful application of this framework in terms of statistical complexity measures (SCMs) \cite{kowalskiEntropy2011}. One common way of defining SCMs is taking the product of a normalized Shannon entropy and an associated measure of disequilibrium \cite{LopezPLA1995,kowalskiEntropy2011}, which captures specific organizational properties of structure and patterns in a process. Following the vast amount of previous work on this topic  \cite{rossoPRL2007,LopezPLA1995,kowalskiEntropy2011,Lange2013}, in this work we consider a variant of SCMs based on the concept of Jensen-Shannon divergence, which is defined as $\mathcal{C}_{JS}[P] = \mathcal{H}[P] \cdot  Q_{JS}[P, P_e]$, where $\mathcal{H}[P]$ is the normalized Shannon entropy and $Q_{JS}[P, P_e]$ is the so-called disequilibrium function measuring the distance of the given distribution $P$ to the uniform distribution $P_e$. Accordingly, we define $Q_{JS}[P, P_e] = Q_{0} \cdot JS[P, P_e]$ where $JS[P, P_e]$ is the Jensen-Shannon divergence between $P$ and $P_e$ and $Q_{0}$ is a normalization factor. Furthermore, the complexity--entropy plane $\mathcal{H} \times \mathcal{C}_{JS}$ has been intensively used to distinguish chaotic systems from stochastic ones \cite{rossoPRL2007}. The analysis of complexity--entropy planes has already taken advantage from the ordinal pattern based symbolic encoding of time series and has found various applications \cite{rossoPRL2007,kowalskiEntropy2011}. However, in the complexity--entropy analysis, there are some chaotic maps that could be easily confused with random noise because no clearly separable domains of values are available to differentiate those different types of dynamics as reported in \cite{BorgesAMC2019}. So far, it remains a challenging task to disclose a possible non-monotonic relationship between complexity and entropy \cite{MartinPLA2003}.

@@

In this work, we draw upon the recent success of both, complexity--entropy plain analysis and OPTNs, to generalize the traditional SCMs by incorporating pattern transition probabilities. The advantages of this approach are as follows: (i) All SCMs successfully show clear distinction between different types of dynamics, which is demonstrated for time series from numerical models and experimental data. (ii) The generalized SCMs help to capture a consistent relationship between SCMs and chaoticity. (iii) All SCMs are sensitive to dynamical transitions along a bifurcation scenario, including period doubling, band merging, inner and outer crises.

The remainder of this paper is organized as follows: In Section~\ref{sec:methods}, we introduce the methodology employed in our work. We start by discussing two slightly different ways to obtain the transition matrix between ordinal patterns, which provides the basis for defining the corresponding OPTNs (Section~\ref{sec:OPW}). Based on this matrix, we propose in Section~\ref{sec:SCM} to compute SCMs from three different perspectives, including static pattern frequency properties, dynamic pattern transition properties, and a combination of both. In Section~\ref{sec:results1}, we then proceed with describing some selected results for the logistic map as as paradigmatic model system exhibiting various types of periodic and chaotic solutions. We first discuss in some detail one example of a periodic orbit to emphasize the important effects of embedding parameters in Section~\ref{sec:embeddings}. The practical usefulness of SCMs will be further illustrated by studying four exemplary time series from different typical dynamical regimes (Section~\ref{sec:four}). In Section~\ref{sec:plane}, the complexity--entropy planes associated with each SCM will be discussed while the control parameter of the logistic map is systematically increased. In order to characterize the dynamical transitions in this model, we further study the behavior of the SCM values in dependence on the control parameter in Section~\ref{sec:transi} and show that similar results can also be obtained for time-continuous dynamical systems like the R\"ossler oscillator in Section~\ref{sec:cont}. Finally, we show that SCMs successfully distinguish different dynamical regimes in experimental time series from flow data and human ECG measurements in Section~\ref{sec:time}. Some concluding remarks and discussions will be provided in Section~\ref{sec:con}.

@@

Most recent studies employing the concept of permutation entropy $\mathcal{S}_O$ have defined this measure by focusing exclusively on the frequencies of ordinal patterns, which however disregards the transition behavior between subsequent patterns. To emphasize the transition frequencies between any pair of patterns, measures of transitional complexity have been further proposed in Refs.~\cite{zhangSciRep2017,McCullough2017b} to quantify the properties of the associated OPTNs. To this end, we associate each directed link in an OPTN with its transition frequency $w_{ij} = p({\pi_i \to \pi_j})$. Calculating the transition frequencies for each pair of patterns, we obtain a weighted directed network characterized by a weighted adjacency matrix $\mathbf{W} = \{ w_{ij} \}, i, j \in [1, D!]$.

Based on this transition matrix $\mathbf{W}$ of an OPTN, we note that there two slightly different ways to introduce normalization factors. The first option is to use a global normalization to ensure $\sum_{i,j}^{D!} w_{ij} = 1$ \cite{zhangSciRep2017}, which results in a globally normalized transition entropy as discussed below. The second option has been proposed by McCullough {\textit{et al}}~\cite{McCullough2017b} who first considered the local out-link transition frequency from pattern $\pi_i$ to $\pi_j$
\begin{equation} \label{eq:localTp}
p_{\pi_i \to \pi_j} = \left \{ \begin{aligned}
				& 0,  \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{if} \;\;\; \pi_i = \pi_j \\
				& \frac{w_{ij}}{\sum_{j, j \neq i} w_{ij}}, \;\;\;\;\;\; \text{if} \; \pi_i \neq \pi_j.\\
				\end{aligned}
				\right.
\end{equation}
\noindent
Note that the transition frequency of Eq.~\eqref{eq:localTp} is pattern (row-) wise normalized. This case will be referred to as the node-wise out-link normalized transition matrix $\mathbf{W}$ in the following. In this way, it easily captures the heterogeneous behavior of both (static) occurrence frequencies of different patterns and (dynamic) transition frequencies between patterns.

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in the time series of R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. On the other hand, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing the weighted matrix $\mathbf{W}$ while acknowledging that in other cases of stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, we have ensured that our results do not change qualitatively when considering the effects of self-loops.

@@

In the following, we will briefly review the existing ordinal pattern frequency based SCM before proposing two novel ways to define SCMs, which explicitly consider transition frequencies of the resulting OPTN representations for time series.

@@

The permutation entropy of a time series is the Shannon entropy of the distribution of patterns $P = \{p_{\pi_1}, p_{\pi_2},\ldots,p_{\pi_i} \}$, which is defined as
\begin{equation}
\mathcal{S}_{O}[P]= - \sum_{i=1}^{D!} p_{\pi_i} \log p_{\pi_i}. 
\end{equation}
The normalized permutation entropy $\mathcal{H}_O$ hence follows as 
\begin{equation} \label{eq:Ho}
\mathcal{H}_{O}[P] = \mathcal{S}_{O}[P] / \mathcal{S}_{O, max}, 
\end{equation}
where $\mathcal{S}_{O, max} = \log D!$ corresponds to the case of a uniform distribution of the same number of patterns, i.e., $P_e = \{1/D!, \ldots, 1/D!\}$.

In the next step, we use some appropriate statistical concept to measure the distance of the given empirical distribution $P$ to the uniform distribution $P_e$, which is the essential idea behind the class of SCMs considered in this work. Several different distance functions have been considered in the literature, including the Euclidean norm, Wooters' distance, Kullback-Leibler relative entropy and Jensen-Shannon divergence \cite{kowalskiEntropy2011}. In this work, the distance between the distributions $P$ and $P_e$ is characterized by the Jensen-Shannon divergence $Q_{JS}[P] = Q_0 \cdot JS[P, P_e]$ with 
\begin{equation}
JS[P, P_e] = \mathcal{S}_O[(P+P_e)/2] - \frac{1}{2}\mathcal{S}_O[P] - \frac{1}{2}\mathcal{S}_O[P_e], 
\end{equation}
where $Q_0$ is a normalization factor with equals the inverse of the maximal Jensen-Shannon divergence $JS[P, P_e]$, which ensures that $Q_{JS}[P] \in [0, 1]$. This maximal distance is obtained when the distribution $P$ has probability one for one pattern and zero for all other patterns. The general expression for $Q_0$ reads  
\begin{equation} \label{eq:Q0}
Q_0 = \frac{-2}{\left(\frac{N+1}{N}\right) \log (N+1) -2 \log (2N) + \log (N)},
\end{equation}
where $N = D!$ is the total number of possible patterns. For large $D$, we have $Q_0 \approx 1 /  \log 2$.

The function $Q_{JS}[P]$ is different from zero if the system has a kind of preference of certain patterns among the accessible ones, reflecting determinism of the system. Finally, the SCM based on permutation entropy of node (pattern) frequencies is defined as 
\begin{equation}
\mathcal{C}_{O}[P] = Q_{JS}[P] \cdot \mathcal{H}_{O}[P].
\end{equation}
This quantity is static in the sense that it quantifies the amount of information stored by ordinal frequencies in the system and its disequilibrium of the observed parts of its accessible states in comparison to a uniform distribution \cite{LopezPLA1995}.

@@

In this case, the sum of all entries of the transition matrix $\mathbf{W}$ is 1, and the Shannon entropy of transition frequencies between any pair of ordinal patterns is therefore 
\begin{equation}
\begin{split}
\mathcal{S}_{W}[W] &= - \sum_{i=1}^{D!} \sum_{j = 1; j \neq i}^{D!} w_{ij} \log w_{ij}, \\
& = - \sum_{i=1}^{D!} \sum_{j =1; j \neq i}^{D!} p_{\pi_i \to \pi_j} \log p_{\pi_i \to \pi_j}. 
\end{split}
\end{equation}
The normalized entropy is hence defined 
\begin{equation}
\mathcal{H}_{W}[W] = \mathcal{S}_{W}[W] / \mathcal{S}_{W, max}, 
\end{equation}
where $\mathcal{S}_{W, max} = \log D! (D! - 1)$ as any self transitions are excluded. Furthermore, there are $D! (D! - 1)$ possible pattern transitions in the uniformly distribution function $P_e$. Thus, in a full analogy with the node-based SCM, the Jensen-Shannon distance is computed as $Q_{JS}[W] = Q_0 \cdot JS[W, P_e]$ with
\begin{equation}
JS [W, P_e] = \mathcal{S}_{W}[(W + P_e)/2] - \frac{1}{2}\mathcal{S}_{W}[W] - \frac{1}{2}\mathcal{S}_{W}[P_e]. 
\end{equation}
We note that we have the same expression for the normalization factor $Q_0$ as above while replacing $N = D! (D! - 1)$ in Eq.~\eqref{eq:Q0}. This further yields the SCM 
\begin{equation}
\mathcal{C}_{W}[W] = Q_{JS}[W] \cdot \mathcal{H}_{W}[W].
\end{equation}
This quantity is dynamic in the sense that it considers the state transition information stored in the system.

@@

\subsection{Choice of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing OPTNs -- the embedding parameters $D$ and $\tau$. The choice of $D$ and $\tau$ may markedly affect the results for the associated SCMs since $D$ and $\tau$ should compromise the essential time scales of the dynamics in the underlying time series. In common applications of ordinal pattern analysis (e.g., estimation of the permutation entropy $\mathcal{S}_O$), the choice of optimal (or just pragmatic) values of $D$ and $\tau$ depends on the field of application, which requires experience of experts as systematically summarized in \cite{Riedl2013}.

@@

As an illustrative example highlighting the important roles of $D$ and $\tau$ for ordinal pattern based analysis, we consider a period-4 solution of the logistic map 
\begin{equation}
x_{i+1}=rx_i(1-x_i) 
\end{equation}
\noindent
with $r= 3.5$. For the sake of simplicity, we choose $D = \tau $. In this case, Fig.~\ref{fig:embed} demonstrates that the resulting ordinal pattern sequences are significantly affected by the different parameter settings, which further result in different permutation entropy values $\mathcal{H}_O$. For instance, only one unique ordinal pattern is observed when $D = \tau = 4$ (Fig.~\ref{fig:embed}(c,g)) resulting in $\mathcal{H}_O = 0$, which is due to the coverage of exactly one full period by the particular embedding parameters. Other values of $D$ and $\tau$ however yield non-zero entropy values. The results of Fig.~\ref{fig:embed} may raise concern since different choices if $D$ and $\tau$ will change the placement of the system in the complexity--entropy plane. In the special case of the logistic map, we do not have a unique choice of $D$ and $\tau$ for periodic windows when the control parameter $r$ is varied. In such cases, we should focus more on the transitions between different dynamic regimes. In the following, we will show that the results are robust when the control parameter is systematically changed while keeping the same parameter settings of $D$ and $\tau$.

\subsection{Results for four typical dynamic regimes} \label{sec:four}

>>>>

\begin{abstract}
Complex network approaches have been recently emerging as novel and complementary concepts of nonlinear time series analysis which are able to unveil many features that are hidden to more traditional analysis methods. In this work, we focus on one particular approach: the application of ordinal pattern transition networks for characterizing time series data. More specifically, we generalize a traditional statistical complexity measure (SCM) based on permutation entropy by explicitly disclosing heterogeneous frequencies of ordinal pattern transitions. To demonstrate the usefulness of these generalized SCMs, we employ them to characterize different dynamical transitions in the logistic map as a paradigmatic model system, as well as real-world time series of fluid experiments and electrocardiogram recordings. The obtained results for both, artificial and experimental data demonstrate that the consideration of transition frequencies between different ordinal patterns leads to dynamically meaningful estimates of SCMs, which provide prospective tools for the analysis of observational time series. 
\end{abstract}

@@

\begin{quotation}
In the recent decade, the field of nonlinear time series analysis has been undergoing fast developments benefiting from concepts from complex network theory. Along this line of research, ordinal pattern transition networks have been expanding the established concept of ordinal time series analysis and provide new insights into the dynamical organization underlying time series data that complement existing methods like permutation entropy. Permutations based on ordinal patterns are a simple and easy to implement concept that naturally provides statistical complexity measures (SCMs), which in the case of permutation entropy rely on pattern frequencies only. Yet, much additional information can be exploited by including order pattern transition frequencies into the definitions of SCMs -- an idea that however has not been widely developed and applied so far. In this work, we generalize existing permutation based SCMs by means of ordinal pattern transition networks which take into account the pattern transition properties explicitly. The usefulness of our generalizations is demonstrated by using time series of both, model and experimental data. 
\end{quotation}

\section{Introduction}
Complex network approaches \cite{ZouPR2018} have recently emerged as prominent tools for nonlinear time series analysis \cite{Kantz97,Sprott2003} and already found many interesting applications to observational and experimental data from various fields. In this line of research, there exist several major methods that are of particular practical importance, including cycle networks \cite{ZhangPRL2006}, recurrence networks \cite{MarwanPLA2009,Donner2010a}, visibility graphs \cite{Lacasa2008,Nunez2012}, and transition networks \cite{Nicolis2005,MichaelChaos2015}. Depending on the particular questions at hand, each of these complex network approaches exhibit various variants that can be employed for different application purposes. For instance, we may either construct a single network representing a univariate time series, or interacting, multiplex or multi-layer networks for coupled time series. For a recent systematic review, see Ref.~\cite{ZouPR2018}. In this work, we focus on constructing ordinal pattern transition networks (OPTNs) derived from time series data, which present the advantage of an easy implementation and a wide range of existing applications to data from different origins, including correlated stochastic processes, neurophysiological (electroencephalogram, EEG) and human cardiac activity (electrocardiogram, ECG) series \cite{MichaelChaos2015,KulpChaos2016,zhangSciRep2017,McCullough2017b,BorgesAMC2019,Subramaniyam2020}.

@@

Given a one-dimensional time series $\{ x_i\}_{i=1, \ldots, L}$ comprising $L$ observations from a dynamical system, we first reconstruct the corresponding phase space by time delay embedding $\vec{x}_i = [x_i, x_{i+\tau}, \ldots, x_{i+(D-1)\tau}]$ with an embedding dimension $D$ \cite{Takens1981,Kantz97}. Next, we represent each embedding vector $\vec{x}_i$ by the corresponding rank order of its components, which is encoded into a symbol $\pi_{i}$ indicating the corresponding ordinal pattern. Hence, when sliding windows from $i=1$ to $N = L - (D - 1)\tau$ in the embedding space, a symbolic representation $\{\pi_i\}_{i=1, \ldots, N}$ of the trajectory is produced. It is possible to derive information about the dynamics of the underlying system by assessing the probabilities of occurrence of different ordinal patterns. For example, for time series resulting from deterministic dynamical systems, certain patterns may not occur at all \cite{AmigoEPL2007,AmigoEPL2008,ZaninChaos2008,RossoEPJB2012}. More generally, we obtain an empirical probability distribution $P$ whose elements $p_{\pi_{i}}$ are the frequencies associated with the different patterns $\pi_{i}$, $i = 1, \ldots, D!$. Clearly, $P$ provides a significant and feasible way to estimate a characteristic probability distribution function for a given time series, which plays a crucial role in extracting statistical complexity information on the underlying dynamical system \cite{BandtPRL2002,AmigoPTRSA2014}.

The conceptually relatively simple computation of $P$ based on ordinal patterns for the underlying system prompts to another successful application of this framework in terms of statistical complexity measures (SCMs) \cite{kowalskiEntropy2011}. One common way of defining SCMs is taking the product of a normalized Shannon entropy and an associated measure of disequilibrium \cite{LopezPLA1995,kowalskiEntropy2011}, which captures specific organizational properties of structure and patterns in the observed process. Following the vast amount of previous work on this topic  \cite{rossoPRL2007,LopezPLA1995,kowalskiEntropy2011,Lange2013}, we consider here a variant of SCMs based on the concept of Jensen-Shannon divergence, which is defined as $\mathcal{C}_{JS}[P] = \mathcal{H}[P] \cdot  Q_{JS}[P, P_e]$, where $\mathcal{H}[P]$ is the normalized Shannon entropy and $Q_{JS}[P, P_e]$ is the so-called disequilibrium function measuring the distance of the given distribution $P$ to the uniform distribution $P_e$. Accordingly, we define $Q_{JS}[P, P_e] = Q_{0} \cdot JS[P, P_e]$ where $JS[P, P_e]$ is the Jensen-Shannon divergence between $P$ and $P_e$ and $Q_{0}$ is a normalization factor. Furthermore, the complexity--entropy plane $\mathcal{H} \times \mathcal{C}_{JS}$ has been intensively used to distinguish chaotic systems from stochastic ones \cite{rossoPRL2007}. The analysis of complexity--entropy planes has already taken advantage from the ordinal pattern based symbolic encoding of time series and has found various applications \cite{rossoPRL2007,kowalskiEntropy2011}. However, in the classical complexity--entropy analysis, there are some chaotic maps that could be easily confused with random noise because no clearly separable domains of values are available to differentiate those different types of dynamics as reported in Ref.~\cite{BorgesAMC2019}. So far, it remains a challenging task to disclose a possible non-monotonic relationship between complexity and entropy \cite{MartinPLA2003}.

@@

In this work, we draw upon the recent success of both, complexity--entropy plane analysis and OPTNs, to generalize an existing SCM by incorporating pattern transition probabilities. The advantages of this approach, which will be demonstrated in the course of this work, are as follows: (i) The modified SCMs allow for a clear distinction between different types of dynamics, which is demonstrated for time series from numerical models and experimental data. (ii) The generalized SCMs help to capture a consistent relationship between SCMs and chaoticity. (iii) All SCMs are sensitive to dynamical transitions along a bifurcation scenario, including period doubling, band merging, inner and outer crises.

The remainder of this paper is organized as follows: In Section~\ref{sec:methods}, we introduce the methodology employed in our work. We start by discussing two slightly different ways to obtain the transition matrix between ordinal patterns, which provides the basis for defining the corresponding OPTNs (Section~\ref{sec:OPW}). Based on this matrix, we propose in Section~\ref{sec:SCM} to compute SCMs from three different perspectives, including static pattern frequency properties, dynamic pattern transition properties, and a combination of both. In Section~\ref{sec:results1}, we then proceed with describing some selected results for the logistic map as a paradigmatic model system exhibiting various types of periodic and chaotic solutions. We first discuss in some detail one example of a periodic orbit to emphasize the important effects of embedding parameters in Section~\ref{sec:embeddings}. The practical usefulness of SCMs will be further illustrated by studying four exemplary time series from different typical dynamical regimes (Section~\ref{sec:four}). In Section~\ref{sec:plane}, the complexity--entropy planes associated with each SCM will be discussed as the control parameter of the logistic map is systematically varied. In order to characterize the dynamical transitions in this model, we further study the behavior of the SCM values in dependence on the control parameter in Section~\ref{sec:transi} and show that similar results can also be obtained for time-continuous dynamical systems like the R\"ossler oscillator in Section~\ref{sec:cont}. Finally, we show that SCMs successfully distinguish different dynamical regimes in experimental time series from flow data and human ECG measurements in Section~\ref{sec:time}. Some concluding remarks and discussions will be provided in Section~\ref{sec:con}.

@@

Most recent studies employing the concept of permutation entropy $\mathcal{S}_O$ have defined this measure by focusing exclusively on the frequencies of ordinal patterns, which however disregards the transition behavior between subsequent patterns. To emphasize on the transition frequencies between any pair of patterns, measures of transitional complexity have been further proposed in Refs.~\cite{zhangSciRep2017,McCullough2017b} to quantify the properties of the associated OPTNs. To this end, we associate each directed link in an OPTN with a transition frequency $w_{ij} = p_{\pi_i \to \pi_j}$. Calculating the transition frequencies for each pair of patterns, we hence obtain a weighted and directed network characterized by a weighted adjacency matrix $\mathbf{W} = \{ w_{ij} \}$ with $i, j \in \{1,\ldots, D!\}$.

Based on this transition matrix $\mathbf{W}$ of an OPTN, we note that there two slightly different ways to introduce normalization factors. The first option \cite{zhangSciRep2017} is to use a global normalization to ensure that $\sum_{i,j}^{D!} w_{ij} = 1$, which results in a globally normalized transition entropy as discussed below. The second option has been proposed by McCullough {\textit{et al.}}~\cite{McCullough2017b} who first considered the local out-link transition frequency from pattern $\pi_i$ to $\pi_j$
\begin{equation} \label{eq:localTp}
p_{\pi_i \to \pi_j} = \left \{ \begin{aligned}
				& 0,  \;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\;\; \text{if} \;\;\; \pi_i = \pi_j \\
				& \frac{w_{ij}}{\sum_{j, j \neq i} w_{ij}}, \;\;\;\;\;\; \text{if} \; \pi_i \neq \pi_j.\\
				\end{aligned}
				\right.
\end{equation}
\noindent
Note that the transition frequency of Eq.~\eqref{eq:localTp} is pattern (row) wise normalized. This case will be referred to as the node-wise out-link normalized transition matrix $\mathbf{W}$ in the following. In this way, it easily captures the heterogeneous behavior of both, (static) occurrence frequencies of different patterns and (dynamic) transition frequencies between patterns.

In addition to the proper normalization, we need to consider self-loops that may affect the numerical estimation of $\mathbf{W}$ especially for time series from continuous systems with a given sampling rate. For example, it has been demonstrated in Ref.~\cite{zhangSciRep2017} that there are about $99\%$ self-loops in time series of the R\"ossler system when integrated with a step size $h = 0.01$, while the about $1\%$ non-self-loops are hidden in $\mathbf{W}$. From this example, one may easily see that self-loops are related to the serial correlation of the time series under study \cite{BorgesAMC2019}. On the other hand, neglecting self-loops in $\mathbf{W}$ emphasizes the transition behavior between different patterns, which has therefore been adopted in most existing studies of complex networks for computational simplicity and theoretical concerns \cite{CostaADPhy2007}. Accordingly, we also remove self-loops in the present study before computing $\mathbf{W}$ while acknowledging that in other cases of stochastic processes and discrete systems, self-loops have been included in the analysis showing some interesting results \cite{BorgesAMC2019}. Arguably, whether to remove or to consider self-loops depends on the particular process under study. In any of the applications discussed in the rest of this paper, we have ensured that our results do not change qualitatively when including self-loops.

@@

In the following, we briefly review the existing ordinal pattern frequency based SCM before proposing two novel ways to define SCMs, which explicitly consider transition frequencies of the resulting OPTN representations for time series.

@@

The permutation entropy of a time series is the Shannon entropy of the distribution of patterns $P = \{p_{\pi_1}, p_{\pi_2},\ldots,p_{\pi_{D!}} \}$, which is defined as
\begin{equation}
\mathcal{S}_{O}[P]= - \sum_{i=1}^{D!} p_{\pi_i} \log p_{\pi_i}. 
\end{equation}
The normalized permutation entropy $\mathcal{H}_O$ hence follows as 
\begin{equation} \label{eq:Ho}
\mathcal{H}_{O}[P] = \mathcal{S}_{O}[P] / \mathcal{S}_{O, max}, 
\end{equation}
where $\mathcal{S}_{O, max} = \log D!$ corresponds to the case of a uniform distribution of the same number of patterns, i.e., $P_e = \{1/D!, \ldots, 1/D!\}$.

In the next step, we use some appropriate statistical concept to measure the distance of the empirical distribution $P$ to a uniform distribution $P_e$, which is the essential idea behind the class of SCMs considered in this work. Several different distance functions have been employed in the literature, including the Euclidean norm, Wooters' distance, Kullback-Leibler relative entropy and Jensen-Shannon divergence \cite{kowalskiEntropy2011}. In this work, the distance between the distributions $P$ and $P_e$ is characterized by the Jensen-Shannon divergence $Q_{JS}[P] = Q_0 \cdot JS[P, P_e]$ with 
\begin{equation}
JS[P, P_e] = \mathcal{S}_O[(P+P_e)/2] - \frac{1}{2}\mathcal{S}_O[P] - \frac{1}{2}\mathcal{S}_O[P_e], 
\end{equation}
where $Q_0$ is a normalization factor with equals the inverse of the maximal Jensen-Shannon divergence $JS[P, P_e]$, which ensures that $Q_{JS}[P] \in [0, 1]$. This maximal distance is obtained when the distribution $P$ has probability one for one pattern and zero for all other patterns. The general expression for $Q_0$ reads  
\begin{equation} \label{eq:Q0}
Q_0 = \frac{-2}{\left(\frac{N+1}{N}\right) \log (N+1) -2 \log (2N) + \log (N)},
\end{equation}
where $N = D!$ is the total number of possible patterns. For large $D$, we have $Q_0 \approx 1 /  \log 2$.

The function $Q_{JS}[P]$ is different from zero if the system has a kind of preference of certain patterns among the accessible ones, reflecting determinism of the system. Finally, the SCM based on permutation entropy estimated from the empirical node (pattern) frequencies is defined as 
\begin{equation}
\mathcal{C}_{O}[P] = Q_{JS}[P] \cdot \mathcal{H}_{O}[P].
\end{equation}
This quantity is static in the sense that it quantifies the amount of information stored by ordinal pattern frequencies in the system and its disequilibrium of the observed parts of its accessible states in comparison to a uniform distribution \cite{LopezPLA1995}.

@@

In this case, the sum of all entries of the transition matrix $\mathbf{W}$ is 1, and the Shannon entropy of transition frequencies between any pair of ordinal patterns is therefore 
\begin{equation}
\begin{split}
\mathcal{S}_{W}[W] &= - \sum_{i=1}^{D!} \sum_{j = 1; j \neq i}^{D!} w_{ij} \log w_{ij}, \\
& = - \sum_{i=1}^{D!} \sum_{j =1; j \neq i}^{D!} p_{\pi_i \to \pi_j} \log p_{\pi_i \to \pi_j}. 
\end{split}
\end{equation}
The normalized entropy is hence defined 
\begin{equation}
\mathcal{H}_{W}[W] = \mathcal{S}_{W}[W] / \mathcal{S}_{W, max}, 
\end{equation}
where $\mathcal{S}_{W, max} = \log D! (D! - 1)$ as any self transitions are excluded. Furthermore, there are $D! (D! - 1)$ possible pattern transitions in the uniform distribution function $P_e$. Thus, in a full analogy with the permutation entropy based SCM, the Jensen-Shannon divergence is computed as $Q_{JS}[W] = Q_0 \cdot JS[W, P_e]$ with
\begin{equation}
JS [W, P_e] = \mathcal{S}_{W}[(W + P_e)/2] - \frac{1}{2}\mathcal{S}_{W}[W] - \frac{1}{2}\mathcal{S}_{W}[P_e]. 
\end{equation}
We note that we have the same expression for the normalization factor $Q_0$ as above when setting $N = D! (D! - 1)$ in Eq.~\eqref{eq:Q0}. This further yields the SCM 
\begin{equation}
\mathcal{C}_{W}[W] = Q_{JS}[W] \cdot \mathcal{H}_{W}[W].
\end{equation}
This quantity is dynamic in the sense that it considers the state transition information stored in the system.

@@

\subsection{Choice of embedding parameters} \label{sec:embeddings}
There are two important algorithmic parameters in constructing OPTNs -- the embedding parameters $D$ and $\tau$. The choice of $D$ and $\tau$ can markedly affect the results for the associated SCMs since $D$ and $\tau$ should compromise the essential time scales of the dynamics in the underlying time series. In common applications of ordinal pattern analysis (e.g., the estimation of the permutation entropy $\mathcal{S}_O$), the choice of optimal (or just pragmatic) values of $D$ and $\tau$ depends on the field of application as systematically summarized in Ref.~\cite{Riedl2013}.

@@

As an illustrative example highlighting the important roles of $D$ and $\tau$ for ordinal pattern based analysis, we consider a period-4 solution of the logistic map 
\begin{equation}
x_{i+1}=rx_i(1-x_i) 
\end{equation}
\noindent
with $r= 3.5$. For the sake of simplicity, we choose $D = \tau $. In this case, Fig.~\ref{fig:embed} demonstrates that the resulting ordinal pattern sequences are significantly affected by the different parameter settings, which further result in different permutation entropy values $\mathcal{H}_O$. For instance, only one unique ordinal pattern is observed when $D = \tau = 4$ (Fig.~\ref{fig:embed}(c,g)) resulting in $\mathcal{H}_O = 0$, which is due to the coverage of exactly one full period by the particular embedding parameters. Other values of $D$ and $\tau$ however yield non-zero entropy values. The results of Fig.~\ref{fig:embed} may raise concern since different choices of $D$ and $\tau$ will change the placement of the system in the complexity--entropy plane. In the special case of the logistic map, we do not have a unique choice of $D$ and $\tau$ for periodic windows when the control parameter $r$ is varied. In such cases, we should focus more on the transitions between different dynamical regimes. In the following, we will show that the results are robust when the control parameter is systematically changed while keeping the same embedding parameters.

\subsection{Results for four typical dynamical regimes} \label{sec:four}

We further illustrate the potential of the proposed SCMs to characterize the logistic map with the control parameter $r$ varying in the range $[3.5, 4]$ with a step size of $\Delta r = 0.001$. In this range of $r$, various dynamical regimes and transitions between them can be found, for instance, period doubling cascades, band merging points, inner and outer crises, and intermittency \cite{Kantz97}, which has made this system serving as a paradigmatic model for assessing nonlinear time series analysis methods \cite{MarwanPLA2009}.
